Modelling the emergence of whisker barrels

### 2019-12-16 paper/briefpaper/brief.tex
Changes to figure and caption with Stuart

%\title{Self-organization of whisker barrels in simulation requires competition between thalamocortical axons}
%\title{Simulated barrel cortex self-organizes using information carried by thalamocortical axons}
\title{Information carried by thalamocortical axons enables self-organization of the barrel field}
% While writing, don't stop for errors
%\nonstopmode

@@

% Please give the surname of the lead author for the running footer
\leadauthor{James}

% Please add a significance statement to explain the relevance of your work
\significancestatement{The mystacial whiskers of rodents are a complex,
  spatially ordered set of structures whose pattern is repeated in several
  locations in the brain. We contribute a self-organising mechanism by which
  the pattern can be recreated with minimal genetic pre-determination at each
  location; only two pairs of axon guidance gradients are required in each
  tissue. The pattern's togography is carried by the axons which grow from one
  tissue to another and interact variably (depending upon which whisker they
  are associated with) with the guidance gradients in the target tissue whilst
  competing with each other for space. We demonstrate that this mechanism
  recreates a realistic whisker barrel field and also accounts for the results
  of well-known gene misexpression experiments.}

@@

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Barrel cortex $|$ Self-organization $|$ Somatotopy $|$ Center-point $|$ Pattern}

\begin{abstract}
Understanding how genetic information and spontaneous pattern formation
together give rise to the whisker barrels is an important challenge for
developmental neurobiology. The barrel patterning that coalesces in layer IV
has been described formally as a Voronoi tesselation, which is indicative of
an underlying developmental process based on local competitive
interactions. If pattern formation relies only on such intrinsic
self-organising processes, how can the somatotopic order at the periphery be
maintained in the projections from brainstem to thalamus and from thalamus to
cortex?
%
To investigate, we began by extending a existing computational model of axon
branching and synaptogenesis in thalamocortical projections under the
influence of molecular signalling gradients.  We found that, although these
processes alone are incapable of generating patterns of cortical fields that
are more intricate than the patterns already present in the signalling
gradients, the addition of competition between the branches of axons that
originate from different thalamic barreloids allowed a full barrel-field
pattern to emerge in the presence of only two gradients.
%
%However, a
%full barrel-field pattern emerges under the guidance of as few as four
%signalling gradients when the model is extended to introduce competition
%between the branches of axons that originate from different thalamic
%barreloids.
%
%We show that the key requirements for the emergence of realistic
%(Voronoi tesselation) patterning are i) at each cortical location
%thalamocortical projections compete for a limited number of available
%connections, ii) at each location the axonal branching rate of each
%thalamocortical projection is reduced by the total branching density of the
%other projections, and iii) the density of axon branching for each
%thalamocortical projection is conserved.
%
We demonstrate how a purely
self-organising system can initialise a pattern which can then be faithfully
replicated in multiple neural structures.
%We show that this minimal model of
%barrel self-organisation provides a parsimonious account of data from a range
%of classic studies in which intrinsic and extrinsic factors have been
%experimentally manipulated.
\end{abstract}

@@

% Blue for comments
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
% Divergence symbol (Del dot Del)
\newcommand{\dvrg}{\nabla\vcdot\nabla}
% Emphasis shortcut
\newcommand{\e}{\emph}
% Bold shortcut
\newcommand{\bol}{\textbf}
% Math bold
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\makeatother

@@

\dropcap{S}patial patterns in neural connectivity provide clues about the
constraints under which brains evolve and develop.
%Whether or not patterning
%governs information processing, models of the development of information
%processing circuits can be validated in terms of their ability to account for
%biological patterning \citep{purves_iterated_1992}.
%%see also \citealp{wilson_what_2015,bednar_cortical_2016}
One of the most
distinctive patterns of neural connectivity is found in the rodent
barrel cortex \cite{woolsey_structural_1970}.
%(\citealp{woolsey_pattern_1948,woolsey_structural_1970,welker_structure_1974},
%see \citealp{fox_barrel_2008}).
Understanding how this particular pattern forms might shed light on the
interplay of genetic and extrinsic factors that shape cortical circuits.
%, and on a wealth of
%anatomical, physiological, and behavioural data obtained using this
%model system.

The barrels are large cortical columns, identifiable from birth in layer 4 of
primary somatosensory cortex as dense clusters of thalamocortical axons, which
are enclosed by borders a few neurons thick from postnatal day 3
\citep{erzurumlu_development_2012}. In the plane tangential to the cortical
surface the barrels constitute a map of the arrangement of the whiskers on the
face,
%(see \citealp{yamakado_subdivision_1979})
such that the cells of adjacent barrels respond most strongly and most quickly
to deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. The
barrels reflect subcortical whisker maps comprising cellular aggregates known
as barreloids in the thalamus \cite{van_der_loos_barreloids_1976} and
barrelettes in the brainstem.
%\citep{killackey_pattern_1980}.

%Each whisker and its corresponding barrelette, barreloid, and barrel can be
%identified by its position on a two-dimensional grid.
%\citep{zucker_coding_1969,killackey_pattern_1980,van_der_loos_barreloids_1976,haidarliu_size_2001}.
%
%The barrels are normally up to twice as long in the axis aligned to the grid
%‘rows’ compared to the ‘arcs’, and the rows are visible from day 2
%\citep{rebsam_refinement_2002} whereas the arcs become distinct by day 3
%\citep{erzurumlu_development_2012,rebsam_refinement_2002}.
%In metabotropic glutamate receptor 5 (mGlu5) knockouts the barrel rows develop
%but the arcs do not (\citealp{hannan_plc-1_2001} see also
%\citealp{fox_barrel_2008}), thus distinct processes likely guide development
%of the barrels with respect to the orthogonal grid axes.

Barrel formation requires afferent input, driven by whisker stimulation and by
thalamic calcium waves that propagate via mGlu5-receptor activation
\citep{anton-bolanos_prenatal_2019}.
%(\citealp{anton-bolanos_prenatal_2019}, see also
%\citealp{anton-bolanos_developmental_2018}).
Patterning also depends on a complex network of axon guidance processes, which
includes interactions between guidance molecules like ephrin-A5 and A7
\citep{miller_epha7-ephrin-a5_2006} and between adhesion molecules like
cadherin-6 and 8 \citep{bishop_regulation_2000}, and the
binding of p75 to neurotropins
\citep{bishop_distinct_2002}.
% or both citations?
%\citep{shimogori_fibroblast_2005,bishop_distinct_2002}.
%
%(\citealp{shimogori_fibroblast_2005,bishop_distinct_2002}
%; see also \citealp{dye_lifespan_2011,dye_lifespan_2011-1}).
This network is orchestrated by interactions between morphogens Fgf8 and Fgf17
and transcription factors Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed
in gradients that mark orthogonal axes and can be manipulated to cause the
barrels to stretch, shrink, shift, skew, and even duplicate \cite{assimacopoulos_fibroblast_2012}.
% A longer list of citations:
%\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012,borello_sp8_2014,sahara_sp8_2007,ypsilanti_transcriptional_2016,sur_patterning_2005}.
%
While complex structure in protein expression can be observed for some genes,
such as the molecules Bhlhb5 \citep{joshi_bhlhb5_2008}, a complex gene
expression pattern with the complexity of the barrel field has not, to our
knowledge been reported and so the mechanism by which the barrels form seems
likely to be one in which the structure forms by self-organization.

The boundaries of the barrels form a Voronoi tesselation across the cortical
sheet \citep{senft_mouse_1991}. This patterning suggests that the barreloid
topology may be preserved in the initial projection of the thalamocortical
axons into the cortex, and that a barrel forms by the lateral branching of the
axons from their initial centres of mass, which ceases upon contact with the
axons branching from adjacent centres. According to this model the barrel
pattern could be entirely determined by the distribution of the initial centre
points, which would itself be determined by guidance cues. However, the
assumption of pre-arranged centres is difficult to resolve with the
observation that thalamocortical axons arrive in the cortical plate as an
undifferentiated bundle prior to barreloid formation within the thalamus
\cite{agmon_organized_1993}. % Plus erzurumli, probably

Alternatively, reaction-diffusion dynamics could generate a Voronoi-like
tesselation in the cortex without the need for pre-arranged centres, by
amplifying characteristic modes in the noisy initial distribution of bundled
axon branches, as a net effect of short-range cooperative and longer-range
competitive interactions between cells. According to this model the barrel
pattern would be determined by the relative strength of these interactions and
by the shape of the cortical field boundary. However, reaction-diffusion
dynamics alone are unlikely to be able to account for the precise
co-registration and topographic connectivity between emergent domains in
thalamus and cortex, the non-uniform sizes and spacing of the barrels to form
rows versus arcs, and the influence of gene expression gradients.

As theories of barrel patterning, the centre-point and reaction-diffusion
models are not mutually exclusive. Pre-organised centres in the initial
conditions could bias reaction-diffusion processes to generate the specific
somatotopic arrangement of barrels more reliably, and similarly the mechanisms
by which axons branch laterally from barrel centres may involve a tension
between cooperative and competitive interactions. However, a proof that barrel
patterning can in principle emerge within an undifferentiated bundle of
thalamocortical axons, based only on local interactions between cells, would
remove the need to invoke a global supervisor, i.e., to assume a separate
stage and/or extrinsic mechanism for pre-organising thalamocortical
connections.

We seek to establish whether it is possible for the observed arrangement of
barrels to self-organise in a system governed by reaction-diffusion dynamics,
under the guidance of molecular signalling gradients, and in the absence of
pre-defined centre-points.
%
Our approach is to represent in a computational model the minimal set of
biological assumptions that might lead to the emergence of barrel patterning
in simulation, and, if necessary, to add to this minimal set only assumptions
that can be represented by the local exchange of information between adjacent
cells. The emergence of barrel patterning under such conditions would serve as
an existence proof of the plausibility of the hypothesis that barrel
development is a self-organising process.

@@

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a dynamical
model of cortical arealization capable of generating predictions about the
influence of signalling molecules on the size, shape, and topological
organization of emergent cortical fields. Their model describes competition
between thalamocortical projections for territory along an anterior-posterior
cortical axis. The equations of the model define how the density of
connections, $c(x,t)$, and the density of axon branches, $a(x,t)$, interact at time $t$
at each cortical location $x$, for $N$ thalamocortical projections indexed by
$i$.
%The Voronoi tesselation used to describe barrel patterning is a
%two-dimensional structure, hence we first extended the original model for
%simulation on a two-dimensional cortical sheet, and describe it here as such
%(i.e., substituting $x$ with the two-dimensional position vector $\mb{x}$ and
%substituting terms of the form $\frac{d}{dx}$ with the spatial gradient
%operator $\nabla$).
%
The model was derived from the assumption that the rates at which
thalamocortical synapses form and thalamocortical axons branch are
reciprocally coupled, resulting in expressions for the change in $c_i$ as a
function of $a_i$, and for the change in $a_i$ as a function of
$c_i$. Synaptogenesis is modelled [in two dimensions, so the variables are
$a_i(\mb{x},t)$ \& $c_i(\mb{x},t)$] as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{i'=1}^{N} c_{i'}\right)[a_i]^k.
\end{equation}
%
When the final term evaluates to zero, the density of thalamocortical
connections decays exponentially at a rate given by $\alpha$. This occurs when
the density of connections over all thalamic projections sums to one;
interpreted as a state in which all available cortical synaptic receptors are
occupied. Otherwise, the connection density is assumed to vary non-linearly
($k>1$) with the axon branching density.

Changes in axon branching density are assumed to depend also on the influence
of guidance molecules and on the density of branches at adjacent cortical
locations:
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of a quantity referred to as the `flux of axonal branching'
(the term enclosed by parentheses, called $\mb{J}_i(\mb{x},t)$).
%The central idea in heat diffusion
%models is that the change in temperature (over time) at a given location is
%proportional to the gradient in the temperature (over space) at that
%location. Substituting the branching density for heat, axons from a given
The flux consists of lateral diffusion across the cortical sheet at a rate
given by the diffusion constant $D$, and an interaction term, in which
branching is influenced by gradients in the concentration of $M$ distinct
molecular signals, $\rho(\mb{x})$. The influence of a given signal (indexed by
$j$) on a given thalamic projection (indexed by $i$), is determined by the
parameter $\gamma_{i,j}$ which may be positive or negative such that axons of
a given thalamic projection may branch preferentially towards either higher or
lower concentrations of each signalling molecule.
%Interactions for
%which this value is positive encourage axon branching in the direction in
%which the concentration of the signal is increasing, whereas negative
%interactions encourage branching in the opposite direction. Hence the axons of
%a given thalamic projection are assumed to branch towards either higher or
%lower concentrations of each signalling molecule.

The second term on the right of Eq.~\ref{eq:da} represents an assumption that the
axon branching density decreases when the number of connections being made
increases, and vice versa, i.e., that branches are required to form synaptic
connections. Note that we introduce the term $\chi=0$ here, so that we might
subsequently redefine $\chi$ to introduce competition between thalamic
projections.
%that we will show to be necessary for the emergence of a
%somatotopic pattern.

Two points are important to emphasise: First, notice that the only
difference between thalamic projections are the interaction strengths,
$\gamma_{i,j}$. Any differences between the size, shape, and location of
cortical fields that reliably emerge from the dynamics of the model must be
due to differences in the choice of these values only; Second, the formulation
of the model is at this point entirely local, in the sense that simulating the
dynamics requires no information to be communicated from cortical locations to
any but the immediately adjacent locations.

In addition to the choice of initial conditions (for the fields
$c(\mb{x},t=0)$ and $a(\mb{x},t=0)$) and the axon guidance fields
$\rho(\mb{x})$, the free parameters of the model are the number of
thalamocortical projections, $N$, constants $\alpha$, $\beta$, $k$, and $D$,
and the $N\times M$ interaction strengths, $\gamma$.

The dynamics of the model (Eqs.~\ref{eq:dc} \& \ref{eq:da}) were solved on a
hexagonal lattice of grid elements representing the cortical sheet, under the
assumption that axon branches cannot extend beyond the boundary of the sheet,
i.e., $\mb{J}_i(\mb{x},t) \big\rvert_{boundary} = 0$. We developed software
tools for imposing arbitrary boundary shapes on the lattice. The dynamics were
numerically integrated using fourth-order Runge-Kutta, and using a
{\color{red} finite difference method} to compute the flux term (see
Supplementary Materials).

@@

%\cmnt{If we need to present the results of the 2-D system with $\chi=0$, then
%  we'll do that first, showing a 2-D version of the simulation obtained by
%  running \code{james0 configs/karb.json}}

We first verified that the system given by Eqs.~\ref{eq:dc} \& \ref{eq:da}
described a two dimensional extension of the axon branching model in
\cite{karbowski_model_2004}. We showed that in the presence of three guidance
gradients varying only in one dimension (along the long axis of an elliptical
domain), five thalamocortical axon types, each interacting uniquely with the
three guidance gradients, would segregate into five separate regions. We
investigated a number of configurations of the guidance gradients, varying
their shape, $\rho(\mb{x})$, and number, $M$, and their effect on a varying
number of thalamocortical projections, $N$. In general, we were not able to
produce a complexity of structure in the resulting patterns of the
connections, $c(\mb{x},t\rightarrow\infty)$, which was not present in the
guidance patterns.

\subsection*{Branching competition is required to form somatotopically ordered barrels}

To investigate how the whisker barrel structure might form in the presence of
only monotonically varying morphogen gradients, we turned our attention to the
competition for connections built into the model. There does exist an element
of competition in the original model (Eqs.~\ref{eq:dc} \& \ref{eq:da}). The
term $\left(1 - \sum_{i=1}^{N} c_i\right)$ in Eq.~\ref{eq:dc} ensures that
once the available synaptic receptors in a region have been fully occupied by
one thalamocortical axon type, there is no longer the possibility for other
types to form synapses. This competition turns out to be relatively weak,
because it does not strongly affect the branching density in that region; it
is possible for two or more TC axon types to have roughly equal branching
density in some region, leading to roughly equal numbers of connections
forming for each TC type, up until the `hard' connection limit is reached.

To introduce competition between the \emph{branching} of thalamocortical axon
types, we introduced the new term $\chi$ given by
%
\begin{equation} \label{eq:comp}
\chi(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{i' \ne i}^{N} f(a_{i'}).
\end{equation}
%
with the choice $f(a) = a$, which leads to a reduction in the rate of increase
of branching density for axon type $i$ in regions where the branching density
of \emph{other} axon types is high. The new parameter $\epsilon$ governs the
strength of the branching competition and the factor $1/(N-1)$ ensures that the
strength scales with $N$.

With the addition of the competition term, $\chi$, it is necessary to
introduce a normalisation step, in order to prevent one TC type from
out-competing every other TC type. Normalisation represents the idea that for
a fixed number of afferent thalamocortical axons, the overall number of
branches must have some upper limit; given the finite resources available in
each axon for building new processes. We divisively normalize the branching
density, dividing it at each computational step by the mean branching density
per unit area. It is then multiplied by the branching density per unit area at
time 0, to account for the possibility that there might be more axons afferent
for TC type $1$ cf.~TC type $2$:
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma}
\end{equation}
%
The prime symbols indicate the use of the intermediate value of $a_i$ after it
has been computed according to Eqs.~\ref{eq:dc}--\ref{eq:comp} on the
basis of the system state at time $t-{\delta}t$.

\subsection*{The model can transfer realistic barrel patterns}

In which we show that a realistic barrel pattern in the thalamus can be
recreated in barrel cortex.

\cmnt{The realistic barrel pattern experiment is enhanced by stronger
  branching competition:}

To increase the strength of the competition, we choose an alternative form for
$f(a)$ in Eq.~\ref{eq:comp} which has a geometric form for small $a$: $f(a) = \frac{1}{1 + e^{-m[a]^l}} - \frac{1}{2}$. The parameters $l$ and $m$ control the steepness and shape of the
function.

%%% End of stronger branching competition %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{./MainFig.png}
\end{center}
%\vspace{-0em}
\caption{A. Barreloid centres taken from \cite{haidarliu_size_2001} are shown as
  black circles. It is assumed that there are two linearly increasing,
  orthogonal signalling gradients (molecules A \& B) in the thalamus (side
  graphs), such that the location of each barreloid within the graded thalamic
  signal field can differentiate the efferent axons, with the result that
  axons from each barreloid interact to varying degrees with cortical
  signalling gradients. Coloured arrows show, for each barreloid, the
  magnitude (and sign) of the interaction that its thalamocortical projection
  will have with the signalling gradients in the cortex. The mappings from
  barreloid location to interaction strength are linear. B. }
\label{fig:main}
\end{figure*}

\subsection*{The barrel field reacts to changes in the expression pattern of guidance molecules}

To cover the reaction to changes in the guidance molecules caused by
experiments like those by Shimogori \& Grove.

@@

We set out to test the hypothesis that barrel patterning may emerge via
self-organising processes. We first extended an existing model of neocortical
arealization to simulate pattern formation in two dimensions, inheriting from
this model the assumption that axonal branching and synaptogenesis are
coupled, and that thalamic projections compete for cortical territory under
the influence of molecular signalling gradients. Realistic barrel patterning
emerged only when terms were added to represent the assumption that
thalamocortical projections compete for a limited supply of connections with
cortical cells, that their branching rates are reduced by the branching of
other thalamocortical projections, and that the density of axon branching is
conserved for each projection. The net effect of these local interactions is a
Voronoi tesselation in which localised clusters of thalamic projections become
somatotopically organised on the cortical sheet according to the relative
values of coefficients specifying the rates at which axons are attracted and
repelled by orthogonal pairs of molecular signal gradients. According to this
model, the pattern of barrels is determined by the competitive and cooperative
interactions between cells relative to the shape of the cortical field
boundary. In simulation, we systematically investigated the origin of the
asymmetry between the barrel rows and arcs, and identified {\color{red} XXXXX}
as the most likely candidate for the elongation of barrel shapes along the
rows.

>>>>

\title{Modelling the emergence of whisker barrels}
% While writing, don't stop for errors
%\nonstopmode

@@

\leadauthor{James}

@@

\keywords{Barrel cortex $|$ Self-organization $|$ Somatotopic map $|$ Axon guidance}

\begin{abstract}
Brain development relies on an interplay of genetic specification and self-organisation. Striking examples of this interplay can be found in the rodent somatosensory brainstem, thalamus, and cortex, where the arrangement of the facial whiskers is preserved in the arrangement of cell aggregates to form precise somatotopic maps. We show in simulation how realistic whisker maps can self-organize, by assuming that information is exchanged between adjacent cells only, under the guidance of gene expression gradients. The resulting model provides a simple account of how genetic information can constrain spontaneous pattern formation to faithfully reproduce functional maps in subsequent brain structures.
\end{abstract}

@@

% Blue for comments
\newcommand{\cmnt}[1]{\textcolor{blue}{#1}}
\newcommand{\dvrg}{\nabla\vcdot\nabla}
\newcommand{\e}{\emph}
\newcommand{\bol}{\textbf}
\newcommand{\mb}[1]{\mathbf{#1}}
\makeatletter
\newcommand*\vcdot{\mathpalette\vcdot@{.35}}
\newcommand*\vcdot@[2]{\mathbin{\vcenter{\hbox{\scalebox{#2}{$\m@th#1\bullet$}}}}}
\makeatother

@@

\dropcap{S}patial patterns in neural connectivity provide clues about the constraints under which brains evolve and develop \citep{purves_iterated_1992}. Perhaps the most distinctive pattern can be found in the rodent barrel cortex \cite{woolsey_structural_1970}. The barrels are large cortical columns, identifiable from birth in layer 4 of primary somatosensory cortex as dense clusters of thalamocortical axons, which are enclosed by borders a few neurons thick from postnatal day 3 \citep{erzurumlu_development_2012}. In the plane tangential to the cortical surface the barrels constitute a somatotopic map of the whiskers, with cells of adjacent barrels responding most strongly and quickly to deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. Barrel patterning reflects subcortical whisker maps comprising cell aggregates called barrelettes in the brainstem and barreloids in the thalamus \cite{van_der_loos_barreloids_1976}.

Barrel formation requires afferent input from whisker stimulation and thalamic calcium waves \citep{anton-bolanos_prenatal_2019}, and depends on a complex network of axon guidance molecules such as ephrin-A5 and A7 and adhesion molecules such as cadherin-6 and 8 \citep{miller_epha7-ephrin-a5_2006}.
This network is orchestrated by interactions between morphogens Fgf8 and Fgf17 and transcription factors Emx2, Pax6, Sp8, and Coup-tf1 \citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed in gradients that mark orthogonal axes and can be manipulated to stretch, shrink, shift, and even duplicate barrels \cite{assimacopoulos_fibroblast_2012}.

The barrel boundaries form a Voronoi tessellation across the cortical sheet \citep{senft_mouse_1991}, suggesting that barreloid topology is preserved in the projection of thalamocortical axons into the cortex, and that a barrel forms by lateral axon branching from an initial centre-point that ceases upon contact with axons branching from adjacent centres. But the assumption of pre-arranged centres is difficult to resolve with the observation that  axons arrive in the cortical plate as an undifferentiated bundle, prior to barreloid formation \cite{agmon_organized_1993}.

Alternatively, reaction-diffusion dynamics could generate a Voronoi tessellation without pre-arranged centres, by amplifying characteristic modes in a noisy initial distribution of axon branches, as a net effect of short-range cooperative and longer-range competitive interactions. Accordingly, the barrel pattern would be determined by the relative strength of these interactions and by the shape of the cortical field boundary. However, intrinsic cortical dynamics alone cannot account for the topographic correspondence between thalamic and cortical domains, the irregular sizes and specific arrangement of the barrels in rows and arcs, or the influence of gene expression gradients.

The centre-point and reaction-diffusion models are not mutually exclusive. Pre-organised centres could bias reaction-diffusion processes to generate specific  arrangements more reliably, and mechanisms of lateral axon branching may constitute the tension between cooperation and competition required for self-organization. However, proof that barrel patterning can emerge from an undifferentiated bundle of  axons, based only on local interactions, would show that a separate stage and/or extrinsic mechanism for pre-organising thalamocortical connections need not be assumed. To this end, we ask whether barrel maps can emerge in a system with reaction-diffusion dynamics, under the guidance of signalling gradients, and in the absence of pre-defined centres.

@@

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a reaction-diffusion style model of how extrinsic signalling gradients can constrain the emergence of distinct fields from intrinsic cortical dynamics. Their model defines how the density of connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a 1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by $i$. The model was derived from the assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally coupled. Extending the original 1D model to simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and  model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one, connections decay at rate $\alpha$. Otherwise the connection density increases non-linearly ($k>1$) with the density of axon branching. Axon branching is modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by $\nabla\vcdot$) of the quantity in parentheses, which is referred to as the `flux' of axonal branching. The flux represents diffusion across the cortical sheet, at rate $D$, and the influence of $M$ molecular signalling fields, $\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which may be positive or negative in order that axons may branch in the direction of either higher or lower  concentrations. The second term on the right represents an assumption that axon branching decreases when synaptogenesis increases, and vice versa, i.e., that axon branches are required for synaptic connections to form. Here $\chi_i=0$ is a placeholder.

Software tools for modelling the cortical sheet as a hexagonal lattice with no-flux boundary conditions and an arbitrary boundary shape, for computing the flux using a finite difference method, for numerical integration by the fourth-order Runge-Kutta method, and for reproducing the reported results are available at \url{https://github.com/ABRG-Models/BarrelEmerge}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{../briefpaper/MainFig.png}
\end{center}

\caption{\textbf{C.} Cortical whisker barrels from \cite{zheng_signal_2001}
  with annotations for approximate axes and lines drawn to join the
  interstices into a putative Voronoi diagram.  \textbf{B.}  Top: Stochastic
  initial condition for $a_i$ with $i=13$ (barrel C4). The roll-off towards
  the boundary (thin line) ensures the boundary conditions are
  satisfied. Bottom: Guidance molecules in the simulated cortex. Molecule 1 is
  expressed high to low in an approximately anterior to posterior
  gradient. Molecule 2 is expressed lateral to medial (analogously to ephrin -
  Vanderhaegen et al) \textbf{C.}  Barreloid centres for mystacial whiskers
  taken from \cite{haidarliu_size_2001} are shown as black circles. The dashed
  grey line shows the boundary of the barreloid field. It is assumed that
  there are two linearly increasing, orthogonal signalling molecules in the
  thalamus, such that the location of each barreloid within the graded
  thalamic signal field can differentiate the efferent axons, with the result
  that axons from each barreloid interact to varying degrees with the cortical
  signalling gradients shown in B. Coloured arrows show, for each barreloid,
  the magnitude (and sign) of the interaction that its thalamocortical
  projection will have with each signalling gradient in the cortex. The
  mappings from barreloid location to interaction strength are
  linear. \textbf{D.}  Barrel cortex simulation. Colours indicate
  $\arg\max_i(c_i)$ for 41 is. Parameters:
  $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
  $\delta{t}=0.0001$. \textbf{E.}  The Honda Dirichlet measure, $\delta$, as a
  function of simulation time for the simulations show in D. \textbf{F.}
  Guidance molecules in the simulated Fgf8 misexpression experiment. Molecule
  1 is mirrored over the anterior-posterior axis The simulation indicates that
  the mirroring of the barrel field can occur as a result of the mirroring of
  guidance molecule 1.}
\label{fig:main}
\end{figure*}

First we verified that all results established by \cite{karbowski_model_2004} for a 1D axis could be reproduced using our extension to a 2D cortical sheet. Using an elliptical boundary with $M=3$ offset guidance gradients aligned to the longer axis, $N=5$ thalamocortical projections gave rise to five distinct cortical fields at locations that preserved the topographic ordering defined by the original $\gamma$ values. However, we found that specifying $N$ ordered areas required $M\approx N$ signalling fields. As the number of distinct guidance fields is unlikely to approach the number of barrels, further extensions were required.

The term in parentheses in Eq.~\ref{eq:dc} represents competition between thalamocortical projections for a limited availability of cortical connections. To introduce competition also in terms of axon branching we redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other projections are dense. Divisive normalization keeps the axon branch density bounded on each iteration to the initial density  at $t=0$;

\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma},
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$.

The only differences between thalamic projections are their strengths of interaction with the guidance fields, $\gamma$, hence any reliable differences between emergent cortical fields must be due to differences in these values only. We speculate that a projection interacts with a given ephrin field in the cortical subplate with a strength determined by the concentration of a similar molecule at its thalamic origin, i.e., the putative barreloid center. As such, two orthogonal linear thalamic gradients were defined, from which 41 pairs of $\gamma$ values were sampled, at the coordinates of 41 barreloid centers recreated from \cite{haidarliu_size_2001}.

A cortical boundary enclosing 41 corresponding barrels was traced from a cytochrome oxidase stain from \cite{zheng_signal_2001}, and Eqs.~\ref{eq:dc}--\ref{eq:norm} were solved for $N=41$ projections on the resulting domain, using $M=2$ linear signalling gradients aligned with the anterior-posterior and  medial-lateral axes. From random initial conditions for $a({\bf x},0)$ and $c({\bf x},0)$, a clear Voronoi-like tesselation emerged. A reduction in the Honda delta metric (see \cite{senft_mouse_1991}) confirmed that a `good' Voronoi pattern emerged within $\approx 3000$ iterations (see Figure 1). The sum of squared differences between the real and simulated barrel centres revealed a strong correspondence.

To further investigate the interplay of genetic and intrinsic mechanisms we simulated a seminal barrel duplication experiment \cite{shimogori_fibroblast_2005},  first by solving on a cortical domain comprising two separate, identically shaped cortical boundaries. In this case one distinct cortical field per thalamic projection emerged, within one of the two boundaries only, i.e., no duplication. However, merging the two fields to create an extended mirror-symmetric boundary shape, with an anterior-posterior guidance field reflected at the join, gave rise to two mirror-symmetrical barrel fields comprising NN barrels, i.e., two cortical fields for each thalamic projection (see Figure 2). Together these results suggest that by the misexpression of Fgf8, Shimogori \& Grove \citep{shimogori_fibroblast_2005} effectively created one large barrel field rather than two distinct fields.

@@

The present results suggest that the key requirements for the emergence of realistic barrel patterning are i) at each cortical location thalamocortical projections compete for a limited number of available synaptic connections (Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a given projection is reduced by the density of other projections (Eq.~\ref{eq:comp}), and iii) the branch density of each projection is conserved (Eq.~\ref{eq:norm}).

It is important to emphasize that the formulation of the model is entirely local, insofar as simulation requires no information to be communicated between thalamic projections, or from a given cortical location to any but those immediately adjacent. Hence the simulations demonstrate how a self-organising system, constrained by genetically specified guidance cues, can faithfully reproduce an arrangement of cell aggregates in one neural structure as a topographic map in another.

The model also shows that the shape of the barrels may be primarily determined by the competitive and cooperative interactions between cells relative to the shape of the cortical field boundary. And that somatotopic map formation does not require the pre-specification of centre-points by as yet undetermined additional developmental mechanisms.
